\chapter{Amazing Life-Journey of the Dhati Community}
\chaptermark{Life-journey}
The story of quitting the \textit{Kshatriya} caste and accepting the pen and
weighing-balance by the king's officers is as courageous and amazing as the
life-journey of the Maheshwari community.

The origin of Maheshwari community is the Marwar region but being a desert and
lack of enough rains they had to face droughts year after year. Such a
situation made the life of Maheshwari families very difficult. In such
conditions, life became a challenge for them and they decided to face it with
exceptional courage. They made small groups and went out of their region in
order to search for their bread-butter and employment.

So, some went to Mewad and others to Jaipur via Ajmer. Yet others went to
Bikaner, crossing Jodhpur border to Pokhran, Falaudi, Jaisalmer, Barmer, Sindh,
Kutchch, Jamnagar etc. places.

According to some Historians, families from Jaisalmer migrated and settled in
Gujarat in the 13th century AD. Families from Mewad went ahead towards
Maharashtra via Gujarat, families from Jaipur went towards Delhi, and those of
Bikaner moved to Calcutta after crossing many borders. People from Marwar also
went to Mumbai and other parts of Maharashtra. These migrations continued for a
long time and many families went and settled in Bengal and current Bangladesh
as well. People from Jodhpur region went towards Bihar, Assam and then to
Utkal-Assam and Nagaland. Some of the Maheshwaris from the Purania district of
Bihar started going to the weekly market at Viratnagar (currently Nepalgunj in
Nepal) and got settled there.

One group of Jaisalmer went to Malwa in Central India and Vidarbh via east
Madhya Pradesh's Gondwana (Jabalpur etc.) and another group went and settled in
Uttar Pradesh's Mathura, Aligarh, Kaasgunj, Meerut and Saharanpur. Such was the
journey of the community and went on to spread in a large part of the country,
setting up an example of courage like other Vaishya communities of
Rajasthan.

For the above mentioned expansion, apart from livelihood, the safety and
security was also a reason. Muslim era -- from Allauddin Khilji to Aurangzeb --
from around year 1300 AD to year 1700 AD -- 400 years and Maratha civil war
were also responsible for such migrations.

Today, Maheshwari community has settled not only in India but have crossed
international borders as well. Considering today's transportation and
communication facilities, it is not very much surprising that people migrate to
different countries but imagining how our ancestors used to protect their
families from thieves-dacoits and enemies using those primitive tools gives
goose-bumps to the most courageous of today's people. We can but only imagine
how by foot, camel and ox-carts, they migrated to unknown regions, mixed with
unknown people, adapted their unknown language and customs and demonstrated
great courage. Such people with their self-courage and firm determination
accepted and faced all kinds of adversities and eventually reached on top
successfully.

Today we are tasting the sweet consequences of our ancestors' holy courage. It
is very satisfying that even today we remain firm in different situations and
face various challenges in order to keep the name of our community high and
have a bright future. Fresh instances of this courage is our migrations during
the 1947 partition and then 1971 war and migrations between 1988-92.

Currently, many Maheshwaris live outside of India of which following are the main countries:
\begin{center}
% use packages: array
\begin{tabular}{ll}
America, Canada and Britain & Approx. 150 to 200 families \\ 
Nepal & Approx. 200 to 250 families \\ 
Bangladesh & Approx. 300 to 400 families \\ 
Sindh (Pakistan) & Approx. 600 to 700 families
\end{tabular}
\end{center}

In the evolution of Maheshwari community it is important to note that they
started with 72 clans that have increased because of various reasons and now
they are approximately 80.

Depending upon the contemporary requirements, keeping away from the fame, using
wisdom and farsighted decision-making is still a lesson for the new generation
as much as it is a requirement of the future growth.
