\chapter{Hospitality}
People of Thar were very simple and straightforward. They were kind, truthful
and self-respecting. Always stayed away from show-off and wiles. If some guest
arrives from other town, they would consider it an honour. A saying went like
this in Maheshwaris: \textbf{ganaat mahnje ankh mathe te} and if someone from
the community arrived as a guest, the saying was: \textbf{nyati kere ghare, koi
roti ro bhukhyo kon ahe} etc. ``Guest is god" was the prevalent sentiment in
those days.

Guest arriving from other town would be offered hand to shake and was asked as
\textbf{chak sakrala, ruda bhala, kher khi, raji khushi, hokara, matara},
similarly the guest would also repeat same words. If some close relative or
friend happens to arrive, they would embrace. And they were asked like so:
\textbf{chadhe aayo ahin ke taane; pani pisho, dyan aane} (means have come
riding camel or pulling it? Would you like some water, shall I bring?). They
would also take care of the camel of the guest.

Then, the guests were offered water or water and light snack. They would be
offered betelnuts, cigarettes etc. One room of the house which was called
`otak' whose one door faced street and the other faced inwards would be used
for the guests. With this kind of system, women of home would be able to
maintain their privacy/decorum. In otak and outside, 2-3 weaved cots were
usually kept. People from town would come to visit the guest at evening/night.
They would exchange news and whereabouts. Those who smoked hukkah would take
turns. Opium would also be mixed and would be shared on palms.

In meals, if there are some sweets they were offered, else quickly something
sweet was cooked. Sometimes, \textit{tikli}, ghee and sugar were offered. They
would share the same plate with guest at meal and would insist on offering food
to the guest. If the guest talks about leaving, they would insist on them to
stay 2-4 days more.

If someone had to go to a city, they would prefer to stay with some relative or
at a community lodge. Generally, they would not stay with other people from
community. If the visited a town or village, and if there is no relative, they
would prefer to stay with the community leader in the town.

When it was time for guests to leave, they would offer some food to take and
would also accompany them to some distance.
