\chapter{Migration from Marwar to Thar}
\chaptermark{Migration}
As mentioned in earlier chapter, the Maheshwari community migrated from Jaisalmer to different parts of India between A.D. 1300 and A.D. 1700 because of different causes. Main reasons for these migrations might be Muslim tyranny, Maratha civil war and continuous drought etc.. ( Only those who had to leave their age-old home and country can know the pain and agony they had to undergo!)

While migrating, where to go was a question. A safe and known place would be a natural choice. Royal states of Jaisalmer and Amarkot (Umarkot) were connected through marriage relationships as the following examples shows:
\begin{enumerate}
 \item Jaisalmer's king Chachakdev first married Umarkot's princess of King Roopsingh in the year A.D. 1197.
\item King Lakhansingh married in Umarkot in the year A.D. 1270.
\item King Jaysingh married Jadawkunwar, daughter of Umarkot's Sodha Naharsang Amarsang.
\end{enumerate}

Since the establishment of Jaisalmer in the year V.S.\footnote{Vikram Samvat, a hindu lunar calendar} 1212 (A.D. 1155) until V.S. 1915 (A.D. 1858) the financial officers were Maheshwaris. So Maheshwaris were also main among the service people of the state governance. While in war, handling rations etc., staying together during wedding ceremonies and in business and royal administration, Maheshwaris used to enjoy the positions of ``ghadvai'', ``choudhary'' and ``mun'hata''. So Umarkot was familiar to Maheshwaris.

In those days, rows of camels (caravan) with many camels used to transport the luggage and goods from the Jaisalmer state. Such rows went to Umarkot via Ratokot towards south (Ratokot was a big city in Thar-Parkar district near the Marwar border. This city was destroyed later.). Another route went through rohdi to sakhkhar. So, that was a known route as well. Both these routes were called \textit{Trade Routes}. Depending upon the situation, familiarity and opportunity, people used one of these routes for migrations. These times were approximately between the years A.D. 1736 and A.D. 1755.

In this way, Maheshwaris settled in the state of Sodha's in Umarkot. Those who travelled on a different route settled in places like Bukera, Tando Alahyaar, Tando Aadam, Sehwan, Badin etc. There were around 300 Maheshwari families in Sehwan at that time (there are none today). Maheshwaris living in Sindh used to speak Thari with an influence of the Sindhi language.

Maheshwaris were strictly vegetarian. Even onion and garlic were considered uneatable. So they could not live with the non-vegetarian culture of Sindh. Apart from that, Marwar was a dry region. On the other hand, the Sindhu river basin was not comfortable because of high humidity and mosquitoes. So they marched forward to Thar desert in search of alternatives. Though Thar was similar to Marwar, they liked it because there was no political tension as was prevalant in Marwar. In such conditions they started looking for their relatives and family members in those areas including the Kutchtch region. Some families that came along Sindh and nearby Mithi and Bagal region were called Sindhi. In 1736 AD, when Mian NoorMohammed attacked Umarkot, Sodhas spread across Thar. With them Maheshwaris too settled in different villages in Thar. These families settled in Dahali, Chhod, Bagal, Chhachharo, Nabisar, Chelhar etc.. In 1875 A.D., there was heavy rains in this region and a lot of domesticated animals died in floods. Because of this several Maheshwaris went to the high and dried sandy regions of Thar. Thus, people from same ``Akaah'' (extended family) settled in one place and in the time of political stability started searching for their families and hence the populations of these places increased.

Along with Maheshwaris, other community people like Pushkarna Brahmin, Saraswat, Shrimali Brahmin, Maali, Sonara, Sutar, Darji, Kumhar etc also got settled.

These migrant Maheshwaris were known by the places they migrated. For example people from Kutchtch were called ``Kutchtchi'', people from Thar were called ``Thari or Dhati'' and people from Jamnagar who came from Nagor (in Marwar) were called ``Nagori''.

This way, Maheshwaris got settled and started developing their business and employment. Gradually they built homes and started marriage etc. rituals. In Thar's various villages 16 out of 72 Maheshwari clans (As described earlier) settled as follows:\\
Rathi, Kela (sarada), Kadva, Hadkut, Gigal, Chandak, Bhutada, Baththar, Malhar, Masania (Baheti), Panpaliya, Lohia, Kachoria, Kasumbia, Malpani, Laghad. Kela's are called ``Ghurya'' in Thar and ``Maandan'' in Kutchtch.

Additional to Thar Villages, some villages in Marwar, such as Sundaro, Mahajalar, Khuhadi, Jaysindar, and Lilmu (These villages are in Indian Territory now) were also associated with giving daughters hand in marriage. Whereas other villages of Marwar were only related with general friendship.
