\chapter{The Origin of the Dhati Community}
\chaptermark{Origin}
Authentic historical literature about the origin of the Maheshwari community is
not available. The sole source of information available is in the form of a
book ``Itihas Kalpdrum Maheshwari Kulbhushan'' authored by late Shivkaranji
Darak of Mundwa. Based on the book, following is the description of the origins
of the Maheshwari community:

Suryawanshi King Khadgalsen of Chauhan dynasty was ruling over Khandelanagar
state. He was very kind and just king. People lived happily and peacefully in
his kingdom. He was always worried of not having a son.

One day the king had invited Brahmins and paid great honour to them. The
Brahmins were very happy with the king's courtesy and asked him for a boon. The
king then expressed his desire for a son. Brahmins said, ``if you worship Lord
Shiva, you will be blessed with a very brave and adventurous son, but do not
allow him to go towards the north and take bath in the Surya-kund there until
he turns 16 years old". If the prince respects Brahmins, he will become a great
king else will be reborn in the same kingdom. Hence being blessed by the
Brahmins, the king rewarded them with fine clothes and jewelry and respectfully
saw them off. The king worshipped God Shiva and was blessed with the boon.

King Khadgalsen had 24 queens. After some time, one of the queens, Champawati
gave birth to a baby boy. The king was very happy and named the prince as Sujan
Kunwar. The prince learned horse-riding, weaponry etc. by the age of 7 years.
When he reached the age of 12, enemies were afraid of him. The king was quite
satisfied with his work. He was careful not to let the prince go towards the
North.

Once a Jain sadhu arrived and preached the prince about Jain religion luring him
into anti-Shiva beliefs and showed the faults of Brahmins. At the age of 14,
the prince opposed Shiva and started practicing Jain religion. He
campaigned the Jain religion in East, West and South and banished idol
worshipping. He harassed Brahmins and broke their sacred threads
(\textit{janoi}). He forbade all religious activities including \textit{yajna}
and \textit{hawan} (holy fire). Out of the King's fear, he never went towards
the North direction, but who can stop the destiny.

Once he went towards the North to the Suryakund with his 72 officers. There he
grew angry when he saw 6 Rishis performing a \textit{yajna}. He ordered his
officers to destroy the\textit{yajna} and harassed the rishis. Seeing this the
rishis cursed them to become stone. So, the prince including all his officers
and horses became stone. This news spread very quickly in all directions.

The King and citizens became worried after hearing this news. King Khadgalsen
died of the shock. 16 of his queens became \textit{sati} (self-immolated) with
him. With no protector of the kingdom, neighbouring enemies attacked the state.
They divided the state into many regions and merged them into their own states.

Even as this happened, the prince's widow and 72 officer's widows cried and
went to the rishis. They humbly requested and begged for the lives of their
husbands. Seeing this rishi had mercy upon them. However, they said they are
not capable enough of revert the curse. They advised the ladies to go to a
nearby cave and worship lord Shiva so that the curse can be taken off. All the
ladies went to a cage and religiously meditated for appeasement of lord Shiva.

After some time, lord Shiva and Parvatiji (lord Shiva's consort) came around
the place where the prince and officers were lying in the form of stones.
Parvatiji asked what happened and lord Shiva told the whole history.

At this time the prince's wife and the officer's wives fell on the feet of
Parvatiji and expressed their plight. Seeing this, Parvatiji requested lord
Shiva of taking off the curse. Lord took off the curse and freed them of the
stone-like state giving them a new life. Everybody fell on lord's feet in
reverance.

As the prince became conscious, his mind filled with lust seeing Parvatiji's
beauty. Seeing this Parvatiji cursed the prince like this: ``O evil man! you
will always beg for food and all your coming generations shall beg for food!"
These people were called \textbf{``jaaga''} (bhat) later on.

The 72 officers said: ``O God! Now we do not have a place to live. What should
we do''? So Shivji informed them, that they quit their \textit{Kshatriya} caste
in a previous birth so now they are liable to accept the \textit{Vaishya}
caste. Go to the Suryakund and have a bath. As they bathed, their sword became
pen, sword-case became stick and shields became weighing balance. All officers
became \textit{Vaishya}. As lord Mahesh (aka Shiva) gave them this lesson, they
were called ``\textbf{Maheshwari}'' \textit{Vaishya}.

When rishis came to know that everybody has been freed of curse, they asked
God: ``O God! how will our incomplete \textit{Yajna} will be completed''? Hence God
preached to the officers that now onwards these rishis are your guru and you
accept them as such. God told the rishis that they do not have anything as of
now but when they have some occasion in their home they will give you material
things to the best of their capabilities. You should teach them to follow their
religion. Rishi accepted them as their pupils and each rishi accepted 12
pupils. Following is their description: (1) Parik from Parashar rishi (2) Dadma
from Dadhichi rishi (3) Adigol from Gautam rishi (4) Khandelwal from Kharik
rishi (5) Sukuwal from Sukumarg rishi (6) Saraswat brahmin/purohit from Sarasur
rishi.

After some time of leaving Khandela all settled in Didwana. From these 72
officers, 72 \textbf{nukh} (clans) came into existence and from these
\textbf{nukh}, depending upon the business, \textit{peta-nukh} (sub-clan) came
into existence.

This day was the ninth day of \textit{Jeth-sud} month. This day is celebrated
as \textit{``Mahesh Navami''} by Maheshwaris. Maheshwari community is
progressing continuously.
