\chapter{Religious Culture}
Religion links soul to the supreme. Religion connects the world with the
creator. Religion shows the way to rendezvous with the God.

There are many religions in this world and people observe their religion. Hindu
religion is the religion of Hindus. It is the oldest religion of all--it is a
\textit{sanatan} (eternal) religion and is also known as a vedic religion.

Thar Maheshwaris are Hindus and followed the Hindu religon. Maheshwari
community originated from the God Mahesh. So, Lord Shiwa was worshipped but
Maheshwaris were Vaishnavs. Every town had a temple. Temple is a leading symbol
of our culture and heritage. Eternal religion is incomplete without temples.
Temples are so integrated in Indian life that it is hard to believe any Indian
has not visited one in his/her lifetime. Many \textit{Murlidhar} temples were
built by Maheshwaris. Beautiful marble idols were brought from Jaipur and were
installed in temples. Shri Ganesh and Hanuman idols would also have a place in
such temples. Towns had \textit{Mataji} temples and \textit{Madhi} where people
would go for worship during \textit{Navratri} fastivities. Other Matajis such
as \textit{Sheetala Mata} idols would also be present in such temples.
\textit{Bhojaks} would maintain and take care of temples and were called
\textit{sevaks}. They would take care of temple's cleanliness, God's bathing,
decoration of the idol, prayers etc. Occasionally, people would donate gold and
silver which was taken care of by the sevaks. Musical instruments such as
drums, hand-symbals etc. would also be found in temples. These were used
occasionally during devotional songs and during festivities such as Holi.

There was not much literacy among Maheshwaris in Thar but there was tremendous
belief in almighty. Sanskrit knowledge was negligible but because of Gujarati
knowledge, people used to read Ramayana, Mahabharat, Bhagvadgita and Bhagvat.
Other religious text were also read. Such Gujarati books were mostly published
by the Akhand Anand Publishers in Ahmedabad city of Gujarat. When people
visited Gujarat for business, they would bring these books with them back home.

Women would visit temple almost everyday. They would donate some grains and
money. They offered 16 \textit{Akshats} (unbroken rice grains) and milk to
\textit{Shivalinga}. After bowing to the god in temple, they would hold a small
mirror put in temple towards god and then towards themselves. From a small bowl
of \textit{tilak}, they would dip a small stick and do a tilak on their
forehead to the tip of their nose. They would circumnavigate the temple. They
would chant devotional songs there. On special festivals such as
\textit{panthevari}, \textit{Holi}, \textit{Janmashtami}, \textit{Shivratri},
etc. related devotional songs would be chanted by community and towns people.

Every 11th and full moon day of the month, woman and some men would observe
fasts. On full moon day meals were taken after seeing and worshipping the moon.
\textit{Ganesh Chaturthi}, \textit{Hanuman Jayanti}, Mondays and Saturdays were
also observed as fasts by some people.

Towns will always have organization of religious stories called \textbf{katha}
in the evening. Usually, elderly women would go to listen to kathas. If there
is a touring playgroup for \textit{ramleela}, men-women would visit without
fail. Occasionally, \textit{Chobas}, \textit{Pandas} and \textit{Gosais} would
visit towns and would receive donations.

Households would organize \textit{Satyanarayan katha} to which neighbors,
relatives and friends were invited to listen. They would listen, and enjoy
sweets and meals in the end.

