\chapter{Tradesmen}
Every town where people from upper caste live, there are people from other
castes who move in. These people are called tradesmen. In Thar, along with
Brahmins, Maheshwaris and Luhanas, such tradesmen lived such as:

\textbf{Gardener}: Would let woman fetch water from their well. Would get bags
of water to home. Would take marriage proposal to groom's place.

Gardener would receive money annually for well water. He would receive some
money for water bags. During weddings he would receive some money and other
gifts from both sides: bride and groom.

\textbf{Barber}: Hindus called them naai and muslims would call them hajaam.
They would come to someone's home and would sit outside on a platform. People
from around the place would come to get their hair dressed or for shaving. They
would receive money or food in return once a year from each family. During the
wedding barber men/women would serve groom/bride. At the end of the wedding
they would receive soiled cloth and shoes from bride/groom. They would also do
minor dressing over boils etc. They would get some gifts during weddings.
Barber women would also aid in childbirth.

\textbf{Cowboy}: Take cattle to feed and also go to needy people's home to milk
their cattle. He would get some money and occasionally gifts and food grains.

\textbf{Potter}: Potters would make and sell clay pots, lamps, pans etc. He
would supply clay made materials during wedding and would receive a good deal
of gifts.

\textbf{Drummer}: A person who would beat drums during special occasions such
as during wedding, or at the time of giving gifts to daughetr after childbirth
or at the time of various other ceremonies. He would get gifts, money and food
grains.

\textbf{Carpenter}: He would also make various wooden items for weddings. He
too would be made happy on these occasions.

Similarly Tailors, Khatri, singers, eunuchs etc would be given their fees and
gifts based on their services. Cash was seldom used.

Trademen were called \textbf{kasabi} in Thar. They would work happily and were
content in what they earned.

